Tool-driven Revolutions in Archaeological Science

### 2019-08-09 analysis/paper/jcaa-paper.Rmd
stash

---
title: "Tool-driven Revolutions in Archaeological Science"
author:
  - author 1
  - author 2
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: journal-of-computer-applications-in-archaeology.csl
abstract: "There is an argument in philosophy of science that revolutions in science are either ideas-driven or tools-driven. We explore this debate in light of recent efforts by many scientific disciplines to embrace methods to improve the reproducibility of their research. One of the most profound changes driven by this concern for reproducibility and transparency is from using data analysis tools dependent on point-and-clicking with a mouse in closed source software, to a set of tools based on writing scripts in open source programming languages and making the scripts openly available. We present bibliometric evidence for this change in ecology and archaeology to test if the adoption of these new tools may be considered a revolution. We identify a citation effect for archaeology articles that use the open source programming language R. We discuss how computational approaches to improving reproducibility and transparency in archaeology are mediated and transformed by the use of R code."
---

@@

# library(caa2018) # Or use devtools::load_all('.', quiet = T) if your code is in script files, rather than as functions in the `/R` diretory

@@

In this paper we investigate recent developments which are facilitating computational archaeology's return to fundamental principles of the scientific method. We ask whether this might be the beginning of a broader change in the field. Revolutions or paradigm-shifts in archaeology have been proposed to occur following major theoretical statements [@Clark_1993; @Härke_2002], but there is an argument among philosophers of science whether revolutions in science are ideas-driven or tools-driven [@Galison_1997; @Dyson_2000]. This argument motivated us to explore how the disciplines of ecology and archaeology have started to embrace tools that improve the reproducibility of research. We focus on one set of tools for this change, based on writing data analysis scripts in free and open source programming languages, exemplified by R, and the practice of sharing of these scripts. We focus on R because our prior observations indicate it is by far the most common scripting language used by ecologists and archaeologists. Bibliometric evidence shows a strong increase in the use of R among ecologists and the beginning of a similar development in archaeology. We evaluate how approaches to improving reproducibility and transparency in archaeology are mediated and transformed by digital approaches and propose these might be a tool-driven change in archaeology. Acknowledging that this is not a simple process, we offer an R-based tool to ease the task of creating a compendium which enables other researchers to reproduce the published results.

# Revolutions in Science: Ideas-driven or Tools-driven?

Philosophers of science disagree whether revolutions in science are ideas-driven or tool-driven. One of the most widely-known models of ideas-driven change in science is Kuhn's [-@Kuhn_1962] effort to describe the history of science by proposing paradigm shifts resulting from revolutionary change in communities of researchers. In brief, this model proposes that the history of science consists of long periods of tradition-bound 'normal science' punctuated by short episodes of 'revolution'. Normal science was described by Kuhn as stagnant, routine day-to-day research focused on 'puzzle solving'. What counts as suitable puzzles, and acceptable solutions to these puzzles, is governed by the norms and procedures of the prevailing paradigm. Revolutions in science occur as punctuations in the equilibria of normal science, when anomalous observations culminate in changes replacing one paradigm with another. In this conversion or speciation process the community adopts a new way of doing science. There has been much debate about how to draw a distinction between so-called normal and revolutionary science. [@watkins1970against; @toulmin1970does; @Casadevalle00158_16]. A notable criticism is that Kuhn's normal science is unscientific, as it describes a situation where critical science had contracted into defensive metaphysics resulting from the domination of a ruling dogma [@popper1970]. This is at odds with a view of science as a continuous state of evolution, with researchers simultaneously employing many styles of thinking and doing science, and continuously confronting theory with evidence and modifying their ideas based on the outcome. In this view, based on the post-World War II era of science, radical discontinuities are rare, and major developments instead emerge from the division and recombination of already mature fields [@andersen2013second]. Given the rarity of radical discontinuities, some biologists have argued that change in biology is better characterised by new ideas (e.g. Mendelian heredity, Darwinian evolution, and molecular genetics) replacing not a former paradigm, but a conceptual vacuum [@wilkins1996are].

In archaeology we have seen claims for paradigm shifts in several areas. The most wide-ranging application is Clark's [-@Clark_1993] comparison of North American archaeology to European archaeology during the 1970s-1990s. His analysis describes a multi-paradigm situation where many research communities (i.e. North American and European) simultaneously operate in different paradigms (which Kuhn noted as a sign of a developing or immature discipline). Clark's account is not a complete Kuhnian analysis because it lacks a revolutionary change event where one paradigm is replaced by another, although he hints that processual archaeology may be considered a radical discontinuity relative to the culture-historical approaches that preceded it [@Clark_1993, pp. 206].

A more thorough treatment of the claim that the appearance of processual archaeology was a paradigm shift is provided by @Meltzer_1979. Meltzer reviews literature arguing for a revolutionary change in archaeology during the 1960s and 1970s, and considers these claims in light of classic Kuhnian revolutionary events such as the replacement of the Ptolemaic system by the Copernican system, and Newtonian dynamics by the new Einsteinian dynamics. He finds that the changes occuring in Anglophone archaeology after the 1960s were incremental, mostly of method, rather than a widespread replacement of one ontological structure by another one, incommensurable or incompatible with the former. Trigger [-@Trigger_2006, pp. 538] came to a similar conclusion at the end of his broad survey of the history of archaeology, that changes throughout archaeology have been mostly additive with only partial replacement.

We also see mentions of paradigm shifts in archaeology in reference to the shift from processual to post-processual archaeology [@Härke_2002; @Koerner_2018], as well as more localised and specific shifts in archaeological practice and thinking. For example, @Snodgrass_2002 claims that among Classical Archaeologists, an increase in research on previously neglected periods of antiquity, such as the Greek Early Iron Age, constitutes a paradigm shift. @McAnany_Rowe_2015 propose that the appearance of community-based participatory models of research among some communities of archaeologists is a paradigm shift, though they concede that it is transformational rather than revolutionary. @harris2012interfacing similarly argues that community crowdsourced geographic knowledge (or volunteered geographic information) could be paradigm shift for archaeological communities. Archaeologists using LIDAR (light detection and ranging) for remote geospatial imaging of cultural landscapes have claimed the application of this technology to archaeology is a paradigm shift [@chase2012geospatial; @howey2016detecting]. @Fuller_2010 has used the term to describe a shift in thinking from the emergence of agriculture as a  'Neolithic Revolution' to a protracted and entangled process, happening several times independently.
Most of these claims for  Kuhnian paradigm shifts differ in meaning, which highlights an important complication with using Kuhn's concept of a paradigm-shift, namely the multiple definitions of a paradigm in his writings.

@@

We propose that Galison's emphasis on tool-driven change, as a complement to Kuhnian concept-driven change, has potential to enhance archaeology's social and scientific relevance and contributions. The centrality of material culture in Galison's view raises the question about whether there is scope for archaeologists to make productive contributions to understanding change in the history of science through an archaeological analysis of scientific instrumentation [cf. @Schiffer_2013]. Furthermore, Galison's analysis of the image/logic contrast in the history of physics invites a similar analysis of the history of archaeology. Can we identify archaeological traditions focused on data collection and analysis using images and image-making devices [cf. @molyneaux2013cultural], in contrast to digital devices? One candidate for this might be the shift from building relative chronologies based on seriation using typologies of visually distinctive artefacts to absolute chronologies based on radiometric dating using computer-controlled instruments. Three revolutions in archaeology have been attributed to radiocarbon dating and associated technologies [@bayliss2009rolling]. @chase2012geospatial have claimed the introduction of radiocarbon dating was a paradigm shift in archaeology, but we are not aware of a Galisonian analysis of this change. A third issue is how we can use a tool-driven approach not just as a framework for understanding change in science, but as a method to predict or generate change in the practice of archaeology?

@@

In this section we explore how a change observed in ecology might be relevant for understanding or even directing the future of archaeology. Biological disciplines have a long tradition of influencing archaeological thought, starting with Oscar Montelius, whose typological method was inspired by Darwin's theory of evolution [@Montelius_1899, 237]. This tradition continues, for example, many spatial statistics currently used by archaeologists are derived from ecology [@keron_use_2015, 7].

@@

@Touchon_McCoy_2016 identify four factors relating to technological change that might explain the changes they observed in the use of statistics in ecology. First, they note that automated data loggers, GPS trackers, remote sensing, and crowd sourcing have greatly increased the rate at which ecologists collect data. Second, increases in desktop and cluster computing power have made complex analytical processes faster and more convenient to compute. Third, the development of free, open source and easily extensible software for data analysis and visualisation, such as R, allows new methods to spread quickly via online fora and social media. Fourth, the publication of several books and papers that have strongly influenced the way many ecologists think about data analysis, such as Burnham and Anderson's [-@burnham2003model] book on model selection and inference.

We focus here on the third factor, free and open source software, because it is the most generic factor and so the most relevant beyond ecology. Archaeology and ecology each have a high diversity of data collection methods, and types of data analyses they conduct. An important similarlity for the two disciplines is working with the field-collected data on a computer to prepare it for publication. R is a widely used free and open source data analysis tool in many research communities [@Tippmann_2015; @Baker_2017; @Thieme_2018]. It is the dominant programming language in archaeology and ecology, so it is a good proxy for the adoption of open source scientific programming languages in these disciplines. To investigate changes in the use of R in archaeology and ecology, we obtained reference lists from a sample of scholarly articles in the Web of Science database and examined patterns in the citation of the R program over time.

@@

We used the 'Cited Reference Search' function provided by the Web of Science online scientific citation indexing service to find journal articles citing R. Although R has been available since the late 1990s [@Thieme_2018], a recommended format for citing the software did not appear until 2004, with the author given as "R Development Core Team". This recommended format for citing the software changed slightly in 2012 when the author was updated to "R Core Team". We searched the Web of Science using '"R DEV COR TEAM" OR "R CORE" OR "R CORE TEAM" OR "R DEVELOPMENT CORE TEAM"' in the CITED AUTHOR field of the Web of Science database. We sorted the results by frequency of citations, and selected the first 1000 items (the maximum allowed by the Web of Science service). These 1000 items represent variations on the recommended format for citing R. We found citations of R in the reference lists of `r prettyNum(how_many_articles_in_top_journals_for_cran_cites, ",")` English-language articles indexed by the Web of Science in the research area of 'Environmental Sciences Ecology'. We then downloaded the bibliographic data and reference lists for each of these articles.

@@

archy_cran_cites_all_areas_top_journals_by_year_to_plot <- 
 archy_cran_cites_all_areas_top_journals_by_year # %>% 
 # filter(SO != "AFRICAN ARCHAEOLOGICAL REVIEW")

all_archaeology_citing_R_over_time <-
  ggplot(archy_cran_cites_all_areas_top_journals_by_year_to_plot,
       aes(PY,
           prop,
           colour = SO)) +
  geom_line(size = 2) +
  geom_point() +
  geom_text_repel(
    data = archy_cran_cites_all_areas_top_journals_by_year_to_plot %>%
      group_by(SO) %>%
      slice(which.max(PY)) ,
    aes(label = SO_n),
    direction = "y",
    size = 6,
    nudge_x = 1,
    hjust = 0) +
   theme(panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(),
         panel.background = element_blank(), 
         axis.line = element_line(colour = "black"),
         axis.title.x = element_text(hjust = 0.125),
        legend.position="none",
        axis.text.x = element_text(angle = 90,
                                   hjust = 1,
                                   vjust = 0.5)) +
  scale_x_continuous(breaks = min_y:2017,
                     limits = c(min_y, max_y)) +
  scale_y_continuous(labels = scales::percent,
                     limits = c(0, 0.2)) +
  xlab("Publication year") +
  ylab("Percentage of articles in that journal") +
  lemon::coord_capped_cart(bottom='right')

@@

jas_cites_r <-
archy_cran_cites_all_areas_top_journals_by_year %>%
  filter(SO == "JOURNAL OF ARCHAEOLOGICAL SCIENCE") %>%
  ggplot(aes(PY,
           prop)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm",
              se = FALSE) +
  stat_poly_eq(aes(label =  paste(..eq.label..,
                                  ..adj.rr.label..,
                                  sep = "~~~~")),
               formula = y~x,
               parse = TRUE,
              label.y = 0.9) +
 stat_fit_glance(method = "lm", 
                  label.y = 0.8,
                  method.args = list(formula = y~x),
                  aes(label = paste("italic(P)*\"-value = \"*", 
                                    signif(..p.value.., digits = 4), sep = "")),
                  parse = TRUE) +
  xlab("year") +
  ylab("Proportion of articles citing R") +
  ggtitle(expression(paste("Articles citing R in ",
                           italic("Journal of Archaeological Science")))) +
  theme_bw(base_size = 10)
```

@@

# combine main plot and subplot
all_archaeology_citing_R_over_time_and_subplot <-
all_archaeology_citing_R_over_time  +
  annotation_custom(grob = ggplotGrob(jas_cites_r),
                          xmin = 2027, xmax = 2040,
                          ymin = 0.10, ymax = 0.20)

@@

The second detail is that most archaeology journals do not show any strong increase in the percentage of article citing R over time. The ecology journals shown in Figure \@ref(fig:ecology-articles-citing-r-plot) show strong upward trends of increasing proportions of articles citing R over time, but we do not see any unambiguous trends in the archaeology articles when considered together. Indeed, in several of the other archaeological journals in our sample, the first citations of R only occurred in the last two years. The sample sizes for the archaeology journals are too small to confidently infer any trend of over time, with the exception of the _Journal of Archaeological Science_ with `r how_many_articles_in_jas` articles. The inset plot in Figure \@ref(fig:archaeology-journals-with-subplot) shows a statistical test of change over time for articles citing R in the _Journal of Archaeological Science_. With a moderate r-squared value and low p-value for the linear model, we conclude there is some evidence for an increase in citations of R over time in this journal. This may show an increasing use of R among archaeologists, especially those working at the intersection of archaeology and the natural sciences. This may hint at the start of a trend like what we see in the ecology journals of a widespread adoption of R.

@@

Figure \@ref(fig:text-mining-article-titles-archaeology-plot) supports our observation that archaeologists using R tend to be doing archaeology that involves the natural and physical sciences. Articles that cite R more often have titles that include terms such as 'lithic', 'experimental', and 'isotope'. On the other hand, articles that do not cite R more often discuss the Roman and Bronze Age periods, and less often reflect a technical specialisation. If there is one technical specialisation that is characteristic of non-citing R articles, it seems to be in ceramics.

@@

@Touchon_McCoy_2016 identified the use of scripting languages as being a factor for easily sharing methods and code, thereby increasing the rate of development of statistical methods. One notable practice we observed during our bibliometric analysis is that among authors of articles citing R, a subset of these authors publicly share the code and data that they used to generate the results presented in their article. We identified `r number_of_reproducible_articles` articles that include R script and data files, either in the supplementary materials or in a trustworthy data repository such as Zenodo, Open Science Framework, Figshare, etc. 'Trustworthy' data repositories are those hosted by a non-commercial organisation, offering persistent identifiers, and have transparent commmitments to long term data management and backup strategies (see @marwick_standard_2018 for more details). Our continuously updated list of articles in this sample is available online here: https://github.com/benmarwick/ctv-archaeology. Figure \@ref(fig:archaeology-reproducible-articles-plot) shows that although this number is small, it is increasing, hinting at the emergence of a new approach to how archaeologists share their research. The author guidelines of this journal support this trend: "In the interests of open scholarship and the reproducibility of results, JCAA strongly encourages all authors to deposit material relating to their publication in a preservation repository" [@authorguidelines_JCAA].

These archaeology articles that include publicly available R code and data are part of a shift in scientific communication that has also been observed in other fields. For example, the journal *Nature Neuroscience* requires authors 'to make the code that supports the generation of key figures in their manuscript available for review' [@Editors_2017]. Although we are not aware of attempts to quantify this change in the same way we have for archaeology, we have observed a number of recent publications that describe and recommend code sharing in statistics [@Baumer2014], genome biology [@markowetz2015five], computational biology [@Sandve2013], biostatistics [@Peng2009], computer science [@peng2011reproducible; @Mitchell2012], applied mathematics [@LeVeque2012], speech science [@Abari2012], political science [@king1995replication; @Dafoe2014], and the social sciences generally [@Miguel2014]. As part of this growing interest in using code for research we also see manifestos aimed at researchers doing any kind of quantitative work [@nosek2015promoting; @Ince2012; @Barnes2010] and articles recommending using R in undergraduate education [@Bray2014; @Eglen2009]. Much of this literature is concerned with identifying and solving problems of irreproducibility in research. These actions come from a renewed interest in basic principles of the scientific method: that once-off results should not be trusted, instead they should be reproducible by other members of the research community [@Stark_2018]. Furthermore, that an article should sufficiently describe the presented research results such that a colleague can fully understand the results and how they were obtained. Many of these authors trace problems of irreproducibility to the increased complexity of computer-based analyses, combined with the limited space to describe them in a journal article, and mouse-driven computer programs where the researcher's analytical decisions are not recorded during the research process, and thus cannot be shared with other researchers.

In our view, merely using R (or similar open source programming languages) to conduct research would not constitute a tool-driven revolution by itself, but publicly sharing the code used for research, as a solution to problems of irreproducibility, is more likely to lead to revolutionary change. We expect that openly shared code will speed the widespread adoption of reproducibility as a core tenet of the scientific process, since it frees researchers from the black box that most mouse-driven computers are, and enables researchers to not only rerun the shared analysis, but to gain access to all parameter settings, empowering them to change these and so properly evaluate, extend and reuse the published results. As open source languages such as R (and Python) are free to use, and most trustworthy repositories are also free (such as Zenodo, Open Science Framework, Figshare, etc.), then even researchers with limited resources anywhere in the world can to contribute equally to the research community by using and sharing code. Using and sharing open source code is thus an important action for reducing inequality in the archaeological research community. It enables participation by researchers who are unable to buy software licenses or access face-to-face training.

@@

```{r  citation-advantage-plot, fig.cap=str_glue('Median citation rates per year for archaeology articles 2010-2017 that cite R (n = {prettyNum(archy_journal_articles_tally[1], ",")}) and articles that do not cite R (n = {prettyNum(archy_journal_articles_tally[2], ",")})')}
knitr::include_graphics(here::here("analysis", "figures", "archy_journal_articles_citations_plot_cap.png"))
```

>>>>

---
title: "Tool-driven Revolutions in Archaeological Science"
author:
  - author 1
  - author 2
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: journal-of-computer-applications-in-archaeology.csl
abstract: "There is an argument in philosophy of science that revolutions in science are either ideas-driven or tools-driven. We explore this debate in light of recent efforts by many scientific disciplines to embrace methods to improve the reproducibility of their research. One of the most profound changes driven by this concern for reproducibility and transparency is from analysing data using tools dependent on point-and-clicking with a mouse in closed source software, to a new set of tools based on writing scripts in open source programming languages and making them openly available. We present bibliometric evidence for this change in ecology and in archaeology to test if the adoption of these new tools may be considered a revolution or some other process. We identify a citation effect for papers that use the open source programming language R. We discuss how computational approaches to improving reproducibility and transparency in archaeology are mediated and transformed by the use of R code."
---

@@

@@

In this paper we investigate recent developments which are facilitating computational archaeology's return to fundamental principles of the scientific method and whether this might be the beginning of a broader change. Revolutions or paradigm-shifts in archaeology have been proposed to occur following major theoretical statements [@Clark_1993; @Härke_2002], but there is an argument among philosophers of science whether revolutions in science are ideas-driven or tools-driven [@Galison_1997; @Dyson_2000]. In this light we explore how the disciplines of ecology and archaeology have started to embrace tools that improve the reproducibility of research. We focus on one set of tools for this change which is based on writing scripts in free and open source programming languages, exemplified by R, and the sharing of these scripts. We focus on R because our prior observations indicate it is by far the most common scripting language used by ecologists and archaeologists. Bibliometric evidence shows a strong increase in the use of R among ecologists and the beginning of a similar development in archaeology. We evaluate how approaches to improving reproducibility and transparency in archaeology are mediated and transformed by digital approaches and propose these might be a tool-driven change in archaeology. Acknowledging that this is not a simple process, we offer an R-based tool to ease the task of creating a compendium which enables other researchers to reproduce the published results.

# Revolutions in Science: Ideas-driven or Tool-driven?

Philosophers of science disagree whether revolutions in science are ideas-driven or tool-driven. One of the most widely-known models of ideas-driven change in science is Kuhn's [-@Kuhn_1962] effort to describe the history of science by proposing paradigm shifts resulting from revolutionary change in communities of researchers. In brief, this model proposes that the history of science consists of long periods of tradition-bound 'normal science' punctuated by short episodes of revolution. Normal science was described by Kuhn as stagnant, routine day-to-day research focused on 'puzzle solving'. What counts as suitable puzzles, and acceptable solutions to these puzzles, is governed by the norms and procedures of the prevailing paradigm. Revolutions in science occur as punctuations in the equilibria of normal science, when anomalous observations culminate in changes replacing one paradigm with another in a conversion or speciation process where the community adopts a new way of doing science. There has been much debate about how to draw a distinction between so-called normal and revolutionary science. [@watkins1970against; @toulmin1970does; @Casadevalle00158_16]. A notable criticism is that Kuhn's normal science is unscientific, a situation where critical science had contracted into defensive metaphysics resulting from the domination of a ruling dogma [@popper1970]. This is at odds with a view of science as a continuous state of revolution, with researchers simultaneously employing many styles of thinking and doing science, and continuously confronting theory with evidence and modifying their ideas based on the outcome. In this view, based on the post-World War II era of science, radical discontinuities are rare, and major developments instead emerge from the division and recombination of already mature fields [@andersen2013second]. Given the rarity of radical discontinuities, some biologists have argued that change in biology is better characterised by new ideas (e.g. Mendelian heredity, Darwinian evolution, and molecular genetics) replacing not a former paradigm but a conceptual vacuum [@wilkins1996are].

In archaeology we have seen claims for paradigm shifts in several areas. The most wide-ranging application is Clark's [-@Clark_1993] comparison of North American archaeology to European archaeology during the 1970s-90s. Clark, similar to Binford and Sabloff [-@binford1982paradigms], identified the North American archaeological paradigm as rooted in anthropology, founded on post-positivist biases, and distinctive with its critical-realist ontology, a modified-objectivist epistemology, and an experimental-manipulative method. In contrast, the European archaeological paradigm is described a kind of history, mostly in a strict empiricist tradition, and typified by a realist ontology, an inductivist epistemology, and by an observational method. Clark's analysis describes a multi-paradigm situation where many research communities (i.e. North American and European) simultaneously operate in different paradigms (which Kuhn noted as a sign of a developing or immature discipline). Clark's account is not a complete Kuhnian analysis because it lacks a revolutionary change event where one paradigm is replaced by another, although he hints that processual archaeology may be considered a radical discontinuity relative to the culture-historical approaches that preceded it [@Clark_1993, pp. 206].

A more thorough treatment of the claim that processual archaeology was a paradigm shift is provided by @Meltzer_1979. Meltzer reviews literature arguing for a revolutionary change in archaeology during the 1960s and 1970s, and considers these claims in light of classic Kuhnian revolutionary events such as the replacement of the Ptolemaic system by the Copernican system, and Newtonian dynamics by the new Einsteinian dynamics. He finds that the changes occuring in Anglophone archaeology after the 1960s were incremental, mostly of method, rather than a widespread replacement of one ontological structure by another one, incommensurable or incompatible with the former. Trigger [-@Trigger_2006, pp. 538] came to a similar conclusion at the end of his broad survey of the history of archaeology, that changes throughout archaeology have been mostly additive with only partial replacement.

We also see mentions of paradigm shifts in archaeology in reference to processual to post-processual archaeology [@Härke_2002; @Koerner_2018], as well as more localised and specific shifts in archaeological practice and thinking. For example, @Snodgrass_2002 claims that among Classical Archaeologists, an increase in research on previously neglected periods of antiquity, such as the Greek Early Iron Age, constitutes a paradigm shift. @McAnany_Rowe_2015 propose that the appearance of community-based participatory models of research among some communities of archaeologists is a paradigm shift, though they concede that it is more transformational rather than revolutionary. @harris2012interfacing has similarly argued that community crowdsourced geographic knowledge (or volunteered geographic information) could be paradigm shift for archaeological communities. Archaeologists using LIDAR (light detection and ranging) for remote geospatial imaging of cultural landscapes have claimed the application of this technology to archaeology is a paradigm shift [@chase2012geospatial; @howey2016detecting]. @Fuller_2010 has used the term to describe a shift in thinking about the emergence of agriculture as a parallel, simultaneous event at the start of the Holocene (the Neolithic Revolution), to new concept of the emergence as a protracted and entangled process, happening 20 or so times independently, following different pathways. Most of these paradigms differ in meaning from Kuhn's classical formulation, and this highlights an important complication with using Kuhn's concept of a paradigm-shift, both as a tool to understand the history of science and as a guide to changing it, which is the multiple definitions of a paradigm in Kuhn's writings.

@@

We propose that Galison's emphasis on tool-driven change, as a complement to Kuhnian concept-driven change, has much potential for archaeology. The centrality of material culture in Galison's view raises the question about whether there is scope for archaeologists to make productive contributions to understanding change in the history science through an archaeological analysis of scientific instrumentation [cf. @Schiffer_2013]. Furthermore, Galison's analysis of the image/logic contrast in the history of physics invites a similar analysis of the history of archaeology. Can we identify archaeological traditions focused on data collection and analysis using images and image-making devices [cf. @molyneaux2013cultural], in contrast to digital devices? One candidate for this might be the shift from building relative chronologies based on seriation using typologies of visually distinctive artefacts to absolute chronologies based on radiometric dating using computer-controlled instruments. Three revolutions in archaeology have been attributed to radiocarbon dating and associated technologies [@bayliss2009rolling]. @chase2012geospatial have claimed the introduction of radiocarbon dating was a paradigm shift in archaeology, but we are not aware of a Galisonian analysis of this change. A third issue is how we can use a tool-driven approach not just as a framework for understanding change in science, but as a method to predict or generate change in archaeology?

@@

The question of predicting or generating change in scientific practice via adoption of new tools is intriguing, and in this section we explore how a change observed in ecology might be relevant for understanding or even directing the future of archaeology. Biological disciplines have a long tradition of influencing archaeological thought, starting with Oscar Montelius, whose typological method was inspired by Darwin's theory of evolution [@Montelius_1899, 237]. This tradition continues, for example, many spatial statistics in current use in archaeology are derived from ecology [@keron_use_2015, 7].

@@

@Touchon_McCoy_2016 identify four factors relating to technological change that might explain the changes they observed in the use of statistics in ecology. First, they note that automated data loggers, GPS trackers, remote sensing, and crowd sourcing have greatly increased rate at which ecologists collect data. Second, increases in desktop and cluster computing power has made complex analytical processes faster and more convenient to compute. Third, the development of free, open source and easily extensible software for data analysis and visualisation, such as R, allows new methods to spread quickly via online fora and social media. Fourth, the publication of several books and papers that have strongly influenced the way many ecologists think about data analysis, such as @burnham2003model's book on model selection and inference.

We focus here on the third factor, free and open source software, because it is the most generic factor and so the most relevant beyond ecology. Archaeology and ecology each have a high diversity of data collection methods, and types of data analyses they conduct, with only a small overlap. What they share in common is working with the field-collected data on a computer to prepare it for publication. R is a widely used free and open source data analysis tool in many research communities [@Tippmann_2015; @Baker_2017; @Thieme_2018] and therefore a good proxy for the adoption of open source scientific programming languages in a discipline. To investigate changes in the use of R in archaeology and ecology, we obtained reference lists from a sample of scholarly articles in the Web of Science database and examined patterns in the citation of the R program over time.

@@

We used the 'Cited Reference Search' function provided by the Web of Science online scientific citation indexing service to find journal articles citing R. Although R has been available since the late 1990s [@Thieme_2018], a recommended format for citing the software did not appear until 2004, with the author given as "R Development Core Team". This recommended format for citing the software changed slightly in 2012 when the author was updated to "R Core Team". We searched the Web of Science using '"R DEV COR TEAM" OR "R CORE" OR "R CORE TEAM" OR "R DEVELOPMENT CORE TEAM"' in the CITED AUTHOR field of the Web of Science database. We sorted the results by frequency of citations, and selected the first 1000 items (the maximum allowed by the Web of Science service). These 1000 items represent variations on the recommended format for citing R. We found citations of R in the reference lists of `r prettyNum(how_many_articles_in_top_journals_for_cran_cites, ",")` English-languge articles indexed by the Web of Science in the research area of 'Environmental Sciences Ecology'. We then downloaded the bibliographic data and reference lists for each of these articles.

@@

all_archaeology_citing_R_over_time <-
ggplot(archy_cran_cites_all_areas_top_journals_by_year,
       aes(PY,
           prop,
           colour = SO)) +
  geom_line(size = 2) +
  geom_point() +
  geom_text_repel(
    data = archy_cran_cites_all_areas_top_journals_by_year %>%
      group_by(SO) %>%
      slice(which.max(PY)) ,
    aes(label = SO_n),
    direction = "y",
    size = 5,
    nudge_x = 1,
    hjust = 0) +
  theme_minimal() +
  theme(legend.position="none",
        axis.text.x = element_text(angle = 90,
                                   hjust = 1,
                                   vjust = 0.5)) +
  scale_x_continuous(breaks = min_y:max_y,
                     labels = c(min_y:2017,
                                rep("",
                                    length(2018:max_y))),
                     limits = c(min_y, max_y)) +
  scale_y_continuous(labels = scales::percent,
                     limits = c(0, 0.2)) +
  xlab("Publication year") +
  ylab("Percentage of articles in that journal")

@@

jas_cites_r <-
archy_cran_cites_all_areas_top_journals_by_year %>%
  filter(SO == "JOURNAL OF ARCHAEOLOGICAL SCIENCE") %>%
  ggplot(aes(PY,
           prop)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm",
              se = FALSE) +
  stat_poly_eq(aes(label =  paste(..eq.label..,
                                  ..adj.rr.label..,
                                  sep = "~~~~")),
               formula = y~x,
               parse = TRUE,
               size = 3,
               label.y.npc = 0.7) +
  stat_fit_glance(label.y.npc = 0.7,
                  size = 3,
                  method = "lm",
                  method.args = list(formula = y ~ x),
                  geom = "text",
                  aes(label = paste("p-value: ",
                                    signif(..p.value..,
                                           digits = 4)))) +
  xlab("year") +
  ylab("Proportion of articles citing R") +
  ggtitle(expression(paste("Articles citing R in ",
                           italic("Journal of Archaeological Science")))) +
  theme_bw(base_size = 6)
```

@@

# combine main plot and subplot
all_archaeology_citing_R_over_time_and_subplot <-
all_archaeology_citing_R_over_time  +
  annotation_custom(grob = ggplotGrob(jas_cites_r),
                          xmin = 2030, xmax = 2040,
                          ymin = 0.10, ymax = 0.17)

@@

The second detail is that most archaeology journals do not show any strong increase in the percentage of article citing R over time. The ecology journals shown in Figure \@ref(fig:ecology-citing-r) show strong upward trends of increasing proportions of articles citing R over time, but we do not see any unambiguous trends in the archaeology articles when considered together. Indeed, in several of the other archaeological journals in our sample, the first citations of R only occurred in the last two years. The sample sizes for the archaeology journals are too small to confidently infer any trend of over time, with the exception of the _Journal of Archaeological Science_ with `r how_many_articles_in_jas` articles. The inset plot in Figure \@ref(fig:archaeology-journals-with-subplot) shows a statistical test of change over time for articles citing R in the _Journal of Archaeological Science_. With a moderate r-squared value and low p-value for the linear model, we conclude there is some evidence for an increase in citations of R over time in this journal. This may show an increasing use of R among archaeologists, especially those working at the intersection of archaeology and the natural sciences. This may hint at the start of a trend like what we see in the ecology journals of a widespread adoption of R.

@@

Figure \@ref(fig:text-mining-article-titles-archaeology-plot) supports our observation that archaeologists using R tend to be doing archaeology that involves the natural and physical sciences. Articles that cite R more often have titles that include terms such as 'lithic', 'experimental', and 'isotope'. On the other hand, articles that do not cite R more often discuss the Roman and Bronze Age periods, and less often reflect a technical specialisation. If there is one that is characteristic of non-citing R articles, it seems to be in ceramics.

@@

@Touchon_McCoy_2016 identified the use of scripting languages as being a factor for easily sharing methods and code, thereby increasing the rate of development of statistical methods. One notable practice we observed during our bibliometric analysis is that among authors of articles citing R, a subset of these authors make publicly available the code and data that they used to generate the results presented in their article. We identified `r number_of_reproducible_articles` articles that include R script and data files, either in the supplementary materials or in a trustworthy data repository such as Zenodo, Open Science Framework, Figshare, etc. Our continuously updated list of articles in this sample is available online here: https://github.com/benmarwick/ctv-archaeology. Figure \@ref(fig:archaeology-reproducible-articles-plot) shows that although this number is small, it is increasing, hinting at the emergence of a new approach to how archaeologists share their research. The author guidelines of this journal confirm this trend: "In the interests of open scholarship and the reproducibility of results, JCAA strongly encourages all authors to deposit material relating to their publication in a preservation repository" [@authorguidelines_JCAA].

These archaeology articles that include publicly available R code and data are part of a new approach to scientific communication that has also been observed in other fields. For example, the journal *Nature Neuroscience* requires authors 'to make the code that supports the generation of key figures in their manuscript available for review' [@Editors_2017]. Although we are not aware of attempts to quantify this change in the same way we have for archaeology, we have observed a number of recent publications that describe and recommend code sharing in statistics [@Baumer2014], genome biology [@markowetz2015five], computational biology [@Sandve2013], biostatistics [@Peng2009], computer science [@peng2011reproducible; @Mitchell2012], applied mathematics [@LeVeque2012], speech science [@Abari2012], political science [@king1995replication; @Dafoe2014], and the social sciences generally [@Miguel2014]. As part of this growing interest in using code for research we also see manifestos aimed at researchers doing any kind of quantitative work [@nosek2015promoting; @Ince2012; @Barnes2010] and articles recommending using R in undergraduate education [@Bray2014; @Eglen2009]. Much of this literature is concerned with identifying and solving problems of irreproducibility in research. These actions comes from a renewed interest in basic principles of the scientific method: that once-off results should not be trusted, instead they should be reproducible by other members of the research community [@Stark_2018]. Furthermore, that an article should sufficiently describe the presented research results such that a colleague can fully understand the results and how they were obtained. Many of these authors trace problems of irreproducibility to the increased complexity of computer-based analyses, combined with the limited space to describe them in a journal article, and mouse-driven computer programs where the researcher's analytical decisions are not recorded during the research process, and thus cannot be shared with other researchers.

In our view, merely using R (or similar open source programming languages) to conduct research would not constitute a tool-driven revolution by itself, but publicly sharing the code used for research, as a solution to problems of irreproducibility, is more likely to enable revolutionary change. Openly shared code speeds the widespread adoption of reproducibility as a core tenet of the scientific process, since it frees researchers from the black box that most mouse-driven computers are, and enables researchers to not only rerun the shared analysis, but to gain access to all parameter settings, empowering them to change these and so properly evaluate, extend and reuse the published results. As open source languages such as R (and Python) are free to use, and most trustworthy repositories are also free (such as Zenodo, Open Science Framework, Figshare, etc.), then even researchers with limited resources anywhere in the world can to contribute equally to the research community by using and sharing code. Using and sharing open source code is thus an important action for reducing inequality in the archaeological research community. This is because it enables participation by researchers who are unable to buy software licenses or access face-to-face training.

@@

```{r  fig.cap=str_glue('Median citation rates per year for archaeology articles 2010-2017 that cite R (n = {prettyNum(archy_journal_articles_tally[1], ",")}) and articles that do not cite R (n = {prettyNum(archy_journal_articles_tally[2], ",")})')}
knitr::include_graphics(here::here("analysis", "figures", "archy_journal_articles_citations_plot_cap.png"))
```
